Manuscript figures

### 2024-11-26 Manuscript_figures.Rmd
Updated R figures R Markdown and HTML report

```{r setup, message=FALSE}
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(pagoda2)
library(scHelper) # github.com/rrydbirk/scHelper
library(cacoa)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ComplexHeatmap)
library(destiny)
library(Seurat)
library(harmony)
library(slingshot)
library(circlize)
library(cowplot)
library(sccore)

@@

@@

We provide lists with filtered count matrices per visit. Please note, we re-sequenced some samples that are renamed in the following.

```{r s1load}
cms <- c("vis1_cms_filtered.qs",
         "vis2_cms_filtered.qs",
         "vis3_cms_filtered.qs") %>% 
  lapply(qread, nthreads = 10) %>% 
  mapply(\(x,y) setNames(x, paste0(names(x), "_", y)), x = ., y = c("visit1", "visit2", "visit3")) %>% 
  unlist() %>% 
  .[!names(.) %in% c("Donor_25_poolB_visit3","Donor_29_poolA_visit3")] %>% 
  setNames(gsub("_poolA|_poolB", "", names(.)))

con.major <- qread("con_major.qs", nthreads = 10)

@@

@@

@@

@@

>>>>

```{r setup, message=FALSE}
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(pagoda2)
library(scHelper) # github.com/rrydbirk/scHelper
library(cacoa)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ComplexHeatmap)
library(destiny)
library(Seurat)
library(harmony)
library(slingshot)
library(circlize)
library(cowplot)
library(sccore)
library(corrplot)

@@

## Figure 3D

Load data

```{r, f3d}
## Plot Individual genes from RNA-seq data
Bulk_norm_counts <- read.delim("Bulk_norm_counts.txt", h=T)
colData <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",")
```

TREM2

```{r, f3d2}
gene_of_interest <- "TREM2"

# Subset norm_counts to get counts for the gene of interest
gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
rownames(gene_counts)=gene_counts$hgnc_symbol
gene_counts=gene_counts[4:96]

# Reshape gene_counts into long format and merge with colData
gene_counts_long <- gene_counts %>%
  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")

gene_counts_long <- gene_counts_long %>%
  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData

# Perform a paired ANOVA separately for each gender
anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))

# Extract ANOVA p-values
anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]

# Perform paired Wilcoxon tests for each sex
wilcox_res <- list(
  visit1_vs_visit2_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit2_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  )
)

wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)

# Create the boxplot with significance annotations
plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers
  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
  theme_minimal() +
  geom_signif(
    comparisons = list(
      c("visit_1.F", "visit_2.F"), 
      c("visit_1.F", "visit_3.F"), 
      c("visit_2.F", "visit_3.F"),
      c("visit_1.M", "visit_2.M"), 
      c("visit_1.M", "visit_3.M"), 
      c("visit_2.M", "visit_3.M")
    ),
    map_signif_level = TRUE,
    test = "wilcox.test",
    test.args = list(paired = TRUE)
  ) +
  theme(plot.title = element_text(size = 10))

# Add ANOVA p-values to the plot
plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)

# Display the plot
print(plot)
```

SPP1

```{r, f3d3}
gene_of_interest <- "SPP1"

# Subset norm_counts to get counts for the gene of interest
gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
rownames(gene_counts)=gene_counts$hgnc_symbol
gene_counts=gene_counts[4:96]

# Reshape gene_counts into long format and merge with colData
gene_counts_long <- gene_counts %>%
  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")

gene_counts_long <- gene_counts_long %>%
  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData

# Perform a paired ANOVA separately for each gender
anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))

# Extract ANOVA p-values
anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]

# Perform paired Wilcoxon tests for each sex
wilcox_res <- list(
  visit1_vs_visit2_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit2_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  )
)

wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)

# Create the boxplot with significance annotations
plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers
  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
  theme_minimal() +
  geom_signif(
    comparisons = list(
      c("visit_1.F", "visit_2.F"), 
      c("visit_1.F", "visit_3.F"), 
      c("visit_2.F", "visit_3.F"),
      c("visit_1.M", "visit_2.M"), 
      c("visit_1.M", "visit_3.M"), 
      c("visit_2.M", "visit_3.M")
    ),
    map_signif_level = TRUE,
    test = "wilcox.test",
    test.args = list(paired = TRUE)
  ) +
  theme(plot.title = element_text(size = 10))

# Add ANOVA p-values to the plot
plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)

# Display the plot
print(plot)
```

MMP9

```{r}
gene_of_interest <- "MMP9"

# Subset norm_counts to get counts for the gene of interest
gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
rownames(gene_counts)=gene_counts$hgnc_symbol
gene_counts=gene_counts[4:96]

# Reshape gene_counts into long format and merge with colData
gene_counts_long <- gene_counts %>%
  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")

gene_counts_long <- gene_counts_long %>%
  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData

# Perform a paired ANOVA separately for each gender
anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))

# Extract ANOVA p-values
anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]

# Perform paired Wilcoxon tests for each sex
wilcox_res <- list(
  visit1_vs_visit2_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_F = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit2_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit1_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  ),
  visit2_vs_visit3_M = wilcox.test(
    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
    paired = TRUE
  )
)

wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)

# Create the boxplot with significance annotations
plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers
  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
  theme_minimal() +
  geom_signif(
    comparisons = list(
      c("visit_1.F", "visit_2.F"), 
      c("visit_1.F", "visit_3.F"), 
      c("visit_2.F", "visit_3.F"),
      c("visit_1.M", "visit_2.M"), 
      c("visit_1.M", "visit_3.M"), 
      c("visit_2.M", "visit_3.M")
    ),
    map_signif_level = TRUE,
    test = "wilcox.test",
    test.args = list(paired = TRUE)
  ) +
  theme(plot.title = element_text(size = 10))

# Add ANOVA p-values to the plot
plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)

# Display the plot
print(plot)
```

@@

We provide lists with filtered count matrices per visit.

```{r s1load}
con.major <- qread("con_major.qs", nthreads = 10)

cms <- con.major$samples %>% 
  lget("misc") %>% 
  lget("rawCounts") %>% 
  lapply(Matrix::t)

@@

## Figure S1E

```{r}
WATLAS <- read.delim("watlas_labels_l1_adipose_anno_super.tsv")
rownames(WATLAS)=WATLAS$X
WATLAS=WATLAS[2:21]

# Define the order of columns and rows
column_order <- c("Adipocyte", "FAP", "VEC", "LEC", "SMC","Pericyte", "B.cell", "CD4..T.cell", "CD8..T.cell", "ILC", "NK.cell", "Mast",  
                  "Macrophage",  "Monocyte", "DC", "pDC",
                  "Endometrium", "Mesothelial", "Plasmablast", 
                  "Schwann" )

row_order <- c("Adipocytes", "ASPC", "EC", "lEC", "SMCs", "Pericytes", "B-cells", "T-cells", "Mast cells", "Myeloid")
WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows

corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
         col = rev(COL2("RdBu",200)),
         is.corr = F, 
         tl.col	= "black",
         tl.pos	= "lower",
         outline = "#C0C0C0",
         addgrid.col= NA)
```

@@

## Figure S2D

```{r}
WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
rownames(WATLAS)=WATLAS$X

# Define Vascular cells
column_order <- c("SOX5..VEC",  "BTNL9..VEC", "ACKR1..VEC", "LEC", "SMC", "Pericyte")
row_order <- c("arEC", "cEC", "cEC2","vEC", "lEC", "SMCs", "Pericytes")
WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows

corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
         col = rev(COL2("RdBu",200)),
         is.corr = F, 
         tl.col	= "black",
         tl.pos	= "lower",
         outline = "#C0C0C0",
         addgrid.col= NA)
```

@@

## Figure S3K

```{r}
WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
rownames(WATLAS)=WATLAS$X

# Define MACS
column_order <- c("LYVE1..Macrophage",  "CYP27A1..Macrophage", "LPL..Macrophage")
row_order <- c("ATM", "mono/mac", "Early LAM", "LAM")
WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows

corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0.5, 1),
         col = rev(COL2("RdBu",200)),
         is.corr = F, 
         tl.col	= "black",
         tl.pos	= "lower",
         outline = "#C0C0C0",
         addgrid.col= NA)
```

@@

## Figure S4D

```{r}
FAP_sub <- read.delim("fap.tsv")
rownames(FAP_sub)=FAP_sub$anno
FAP_sub=FAP_sub[2:5]

# Define ASPCs 
column_order <- c("FAP2","FAP1", "FAP4","FAP3")
row_order <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
FAP_ordered <- FAP_sub[, rev(column_order)] # Reorder columns
FAP_ordered <- FAP_ordered[row_order, ] # Reorder rows

corrplot(t(t(FAP_ordered)), method="color", order = , col.lim=c(-1, 1.09),
         col = rev(COL2("RdBu",200)),
         is.corr = F, 
         tl.col	= "black",
         tl.pos	= "lower",
         outline = "#C0C0C0",
         addgrid.col= NA)
```

## Figure S4E

```{r}
WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
rownames(WATLAS)=WATLAS$X

# Define ASPCs 
column_order <- c("DPP4..FAP",  "CXCL14..FAP", "PPARG..FAP", "ICAM1..FAP")
row_order <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows

corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0.4, 1),
         col = rev(COL2("RdBu",200)),
         is.corr = F, 
         tl.col	= "black",
         tl.pos	= "lower",
         outline = "#C0C0C0",
         addgrid.col= NA)
```
